---
title: "Testing the Aassumptions of ANOVAs"
author: "Mattan S. Ben-Shacahr"
date: "`r Sys.Date()`"
show_toc: true
output:
  knitr:::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Testing the Aassumptions of ANOVAs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 90)
knitr::opts_chunk$set(dpi=72)
```

`afex` comes with a set of built-in functions to help in the testing of the assumptions of ANOVA design. Generally speaking, the testable assumptions of ANOVA are^[There is also the assumptions that (a) the model is correctly specified and that (b) errors are independent, but there is no "hard" test for these assumptions.]:

1. ***Homogeneity of Variances***: the variances across all the groups (cells) of between-subject effects are the same. This can be tested with `performance::check_homogeneity()`.  
2. ***Sphericity***: For within-subjects effects, sphericity is the condition where the variances of the differences between all possible pairs of within-subject conditions (i.e., levels of the independent variable) are equal. This can be thought of as a within-subjects' version of the *Homogeneity of Variances* assumption, and can be tested with `performance::check_sphericity()`.   
3. ***Normality of residuals***: The errors used for the estimation of the error term(s) (*MSE*) are normally distributed. This can be inferred using `performance::check_normality()`.

What follows is a brief review of these assumptions and their tests.

```{r, message=FALSE}
library(afex)
library(performance) # for assumption checks
```


## Homogeneity of Variances


This assumption, for between subject-designs, states that the within group errors all share a common variance around the group's mean. This can be tested with Levene's test:


```{r}
data(obk.long, package = "afex")

o1 <- aov_ez("id", "value", obk.long, 
             between = c("treatment", "gender"))

check_homogeneity(o1)
```
These results indicate that homogeneity is not significantly violated.


### What to do when assumption is violated?

ANOVAs are generally robust to "light" heteroscedasticity, but there are various other methods (not available in `afex`) for getting robust error estimates.

Another alternative is to ditch this assumption altogether and use permutation tests (e.g. with [`permuco`](https://cran.r-project.org/package=permuco)) or bootstrapped estimates (e.g. with [`boot`](https://cran.r-project.org/package=boot)).

## Sphericity

```{r}
data("fhch2010", package = "afex")

a1 <- aov_ez("id", "log_rt", fhch2010,
             between = "task", 
             within = c("density", "frequency", "length", "stimulus"))
```

We can use `check_sphericity()` to run Mauchly's test of sphericity:

```{r}
check_sphericity(a1)
```
We can see that both the error terms of the `length:stimulus` and `task:length:stimulus` interactions significantly violate the assumption of sphericity at *p* = 0.021. Note that as `task` is a between-subjects factor, both these interaction terms share the same error term!

### What to do when assumption is violated?

- For ANOVA tables, a correction to the degrees of freedom can be used - `afex` offers both the Greenhouse-Geisser (which is used by default) and the Hyunh-Feldt corrections.  
- For follow-up contrasts with `emmeans`, a multivariate model can be used, which does not assume sphericity.

Both can be set globally with:
```{r, eval = FALSE}
afex_options(
  correction_aov = "GG", # or "HF"
  emmeans_model = "multivariate"
)
```


## Normalicy of Residuals

The normalicy of residuals assumption is concerned with the errors that make up the various error terms in the ANOVA. Although the Shapiro-Wilk test can be used to test for deviation from a normal distribution, [this test tends to have high type-I error rates](https://notstatschat.rbind.io/2019/02/09/what-have-i-got-against-the-shapiro-wilk-test/). Instead, one can visually inspect the residuals using quantile-quantile plots (AKA qq-plots). For example:

```{r}
data("stroop", package = "afex")

stroop1 <- subset(stroop, study == 1)
stroop1 <- na.omit(stroop1)

s1 <- aov_ez("pno", "rt", stroop1,
             within = c("condition", "congruency"))

is_norm <- check_normality(s1)

plot(is_norm)

plot(is_norm, type = "qq")
```
If the residuals were normally distributed, we would see them falling close to the diagonal line, inside the 95% confidence bands around the qq-line.

We can further de-trend the plot, and show not the expected quantile, but the deviation from the expected quantile, which may help reducing visual bias.

```{r}
plot(is_norm, type = "qq", detrend = TRUE)
```

Wow! The deviation from normalicy is now visually much more pronounced!

### What to do when assumption is violated?

As with the assumption of *homogeneity of variances*, we can resort to using permutation tests for ANOVA tables and bootstrap estimates / contrasts.

Another popular solution is to apply a monotonic transformation to the dependent variable. This should not be done lightly, as it *changes* the interpretability of the results (from the observed scale to the transformed scale). Luckily for us, it is common to *log* transform reaction times, which we can easily do^[But note ANOVA no longer tests if any differences between the means is significantly different from 0, but if any ratio between the means is significantly different from 1.]:

```{r}
s2 <- aov_ez("pno", "rt", stroop1,
             transformation = "log",
             within = c("condition", "congruency"))

is_norm <- check_normality(s2)

plot(is_norm, type = "qq", detrend = TRUE)
```

Success - after the transformation, the residuals (on the log scale) do not deviate more than expected from errors sampled from a normal distribution (are mostly contained in the 95%CI bands)!

# References 

<!-- manual -->